|

### 2023-07-19 manuscript/manuscript_1.0.Rmd
Addressed Daniel's comments

---
title: |
  The role of learning in status signalling: a modeling approach
author:
- Andrés E. Quiñones:
    email: "andresqp@gmail.com"
    institute: [andes, unine]
    correspondence: true
- Redouan Bshary:
    institute: unine
- C. Daniel Cadena:
    institute: andes
institute:
- andes: Laboratorio de Biología Evolutiva de Vertebrados, Departamento  de Ciencias Biológicas, Universidad de los Andes, Bogotá, Colombia
- unine: Institute of Biology, University of Neuchâtel, Neuchâtel, Switzerland
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  # word_document: default
  pdf_document:
    latex_engine: xelatex # default  #
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
fig_caption: yes
fontsize: 12pt
header-includes: |
  ```{=latex}
    \usepackage{float}
    \usepackage{unicode-math}
    \usepackage{setspace}\doublespacing
    \usepackage[backref=page]{hyperref}
    \linespread{2}
    \usepackage{lineno}
    \linenumbers
    \floatplacement{figure}{H}
    \newcommand{\beginsupplement}{ \setcounter{table}{0}         \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
  ```
keep_tex: yes
keywords: My keywords
linespread: 2
bibliography: ../SocialMorphCue.bib
spacing: double
style: C:\Users\andre\Zotero\styles\elife.csl
abstract: |
  Adaptive behavioural responses often depend on qualities of the interacting
  partner of an individual. For example, when competing for resources, an individual
  might be better off escalating fights with individuals of lower quality, while
  restraining  from fighting individuals of higher quality. Communication systems
  involving signals of quality allow individuals to reduce uncertainty regarding the
  fighting ability of their partners and make more adaptive behavioral decisions. 
  However, dishonest individuals can destabilize such communications systems. 
  One open question is whether cognitive mechanisms, such as learning, 
  can maintain the honesty of signals, thus, favoring their evolutionary 
  stability. Here we present evolutionary simulations, where individuals 
  can produce a signal proportional to their quality; and learn along their
  lifetime the best response to the signal emitted by their peers. Our simulations
  replicate previous results where the handicap principle mediates the evolution 
  of signals. As in traditional models without learning, in our simulations 
  learning on the receiver side can mediate the evolution of signals
  of quality on the sender side. When the cost of the signal is proportional to the
  quality of the sender populations are only composed of honest signalers. When the
  cost is not proportional to the quality of the signaler tranditional models 
  predict the absense of a signal, in contrast learning facilitates the 
  evolution of a polymorphim where population is composed of both honest and 
  dishonest signalers. We argue that learning can be a general cognitive mechanism 
  role wide range of communication systems and more generally
  behavioral responses."
---

@@

The outcome of interactions with conspecifics individuals is an important 
determinant of fitness in social animals. Irrespective of whether 
such interactions are cooperative or competitive, the actions of 
interacting individuals partly determine 
their reproductive success. However, the best action for an individual 
engaging in a social interaction might vary depending on its 
own condition, and that of individuals with 
whom it interacts. Thus, acting based on information about interacting 
partners is typically adaptive [@quinones_Negotiation_2016], yet
acquiring such information is far from trivial. 
In some cases, interacting partners (signallers) might be 'willing' to 
provide accurate information, but in others it might be in their 
own interest to conceal information [@johnstone_Recognition_1997], 
or to provide deceptive information [@johnstone_Badges_1993]. 
Typically, in a given context, some individuals benefit 
from broadcasting accurate information, whereas others benefit 
from concealing it. Take, for instance, an interaction between two individuals 
where one can help the other. Given some costs and benefits, 
a potential donor is likely to be interested in helping 
related individuals. Therefore, for a relative of the donor, 
broadcasting kinship would be advantageous, whereas concealing the lack 
of kinship would be better for an unrelated individual. 
Similar scenarios may apply to a variety of aspects of social 
life such as finding mates, feeding offspring, or engaging in
dominance relationships or aggressive interactions 
[@bradbury_Principles_2011; @moller_Female_1988; @tibbetts_Socially_2004].

In dominance and aggressive interactions, a crucial piece of
information to guide individual actions is the fighting ability of interacting 
partners, which is sometimes referred to as Resource Holding Potential 
[RHP, Parker -@parker_Assessment_1974], or simply as quality. 
Responsiveness to the quality of an opponent is central 
to communication systems such as those in which individuals 
exhibit badges of status, where a signal conveys quality 
[@johnstone_Badges_1993; @rohwer_Social_1975]. 
Such signals must be arbitrary, meaning the signals could potentially be 
produced by high and low quality individuals alike. 
Generally signals can be evolutionarily stable whenever low quality individuals 
pay a higher fitness cost for carrying the signal such that it is not 
in the interest of low quality individuals 
to fake quality by producing a large badge. Signal costs inversely proportional
to quality can come about as a result of the production costs of the signal, 
in which case the signal works like a handicap [@botero_Evolution_2010; @johnstone_Badges_1993;@grafen_Biological_1990; @zahavi_Mate_1975]. 
Alternatively, the costs can come from 
the aggressive reaction of receivers when the convention of the communication 
is broken [@enquist_Signaling_2010;@tibbetts_Socially_2004]. 
<!-- should this last part of the paragraph stay ?-->
<!-- An alternative system of communication is one where individuals  -->
<!-- gather and store information about each individual they interact with.  -->
<!-- This information should reliably allow each individual to react adaptively  -->
<!-- to the quality difference with their opponent. This type of  -->
<!-- assessment strategy is often referred to as individual  -->
<!-- recognition [@whitfield_Plumage_1986]. Evidently, individual recognition  -->
<!-- is restricted to  certain interaction structures, because individuals  -->
<!-- are limited by cognitive abilities in how many individuals  -->
<!-- they can recognize and track. Thus, individual recognition and  -->
<!-- badges of status have been assumed to be alternative communication  -->
<!-- systems fulfilling the same function, reducing the costs of  -->
<!-- aggression in contests over resources.  -->

@@

A somewhat ignored component of communication systems in the context 
of aggressive interactions are cognitive aspects of the receiver module. 
Theoretical models often assume that communication systems relying on badges of 
status have reaction norms as their mechanistic underpinning 
[@botero_Evolution_2010], with 
individuals using their own quality and the badges of opponents to determine 
whether to aggressively engage in contests. Reaction norms 
allow individuals to respond to the available information without a large
cognitive burden because the computations involved in the reaction 
norm response do not require much memory nor advanced algorithmic processes. 
This contrasts with other information-processing 
mechanisms like individual recognition. With individual recognition, 
where individuals associate cues of their peers to their quality. Because these 
associations must be learnt throughout the life of individuals, 
they have the usual cognitive requirements of associative learning processes. 
An alternative view of badges of status is that individuals learn 
to react to them based on their experiences [@guilford_Receiver_1991], 
which would imply that receivers must learn to associate signals with 
the fighting ability of bearers just as in systems based on individual 
recognition. In cases where badges of status vary quantitatively 
(e.g. in size or intensity), fighting ability may increase monotonically 
with attributes of the badge and would be reinforced by every interaction.
Therefore, in principle, learning in such scenarios would be faster 
than in systems involving individual 
recognition in which the association between signals and their meaning 
varies depending on the interacting partners. In any case, learning may be 
a central cognitive mechanism in both types of communication systems, 
but the role of learning in these contexts has not been thoroughly 
explored in the empirical nor theoretical literature.

@@

Associative learning is a taxonomically widespread cognitive 
mechanism that allows individuals to associate rewards with environmental stimuli 
and thus behave adaptively 
[@heyes_Simple_2012;@macphail_Brain_1982; @staddon_Adaptive_2016; 
@behrens_Associative_2008].
Theory has shown that natural selection favours such associations 
in complex environments where conditions are difficult to predict 
[@dridi_Environmental_2016]. Associative learning is a flexible cognitive 
mechanism whose underpinnings show interspecific variation 
[@enquist_Power_2016; @quinones_Reinforcement_2020]. Associative learning 
is not a single mechanisms, but rather a set of cognitive structures 
and processes that can vary in their scope and complexity. Presumably, 
these structures and processes have been modified 
by natural selection and could provide new explanations for behavioural 
variation. Despite all this, associative learning is 
not often included in the narrative of evolutionary explanations of 
behavioural patterns [@fawcett_Exposing_2013; @kamil_Optimal_1983; 
@mcauliffe_Psychology_2015]. Computational models of evolution 
can play an important role to overcome the lack of integration 
between learning and evolution. Reinforcement learning theory 
encompasses a series of computational methods inspired 
on the psychological and neurological mechanisms of associative 
learning [@sutton_Reinforcement_2018]. This set of algorithms allows the 
implementation of biologically realistic problems, 
capturing the essence of learning processes 
[@frankenhuis_Enriching_2018; @quinones_Reinforcement_2020]. 
Furthermore, these algorithms can be embedded in evolutionary simulations to 
generate theoretical predictions of the effect that learning can 
have in behavioural evolution [@leimar_Learning_2019;@leimar_Effects_2022].

Here we present an evolutionary model where individuals use 
associative learning to develop a tendency to behave aggressively 
(or  peacefully) in the context of competition over resource, 
depending on a quantitative morphological trait 
(i.e.: a badge indicating quality) they 
perceive in their opponents. Over evolutionary time the size of 
the badge evolves, and its dependency on the individual's quality. 
Under this simple set up, individuals can use the badge as a signal of quality. 
We use the model to assess under what conditions 
we expect communication signals to evolve as handicaps or conventions.

@@

We model the evolution of signals indicating individual quality 
in the context of agonistic interactions. 
For simplicity we consider a population of haploid 
individuals with non-overlapping generations. Individuals are born 
every generation with a level of quality ($Q_i$, where $i$ is subscript 
of the population vector of size $N$) given by a number between zero and one, 
which is drawn from a truncated normal distribution 
$N( 0.5,\sigma)$. An individual with quality $0$ has the lowest RHP, 
while an individual with quality $1$ has the highest. As they 
develop, individuals produce a phenotypic signal (badge), 
the size of which depends on their quality according to a reaction norm given by 
$B_i=1/(1+e^{\alpha_i-\beta_i Q_i})$; where $\alpha_i$ and 
$\beta_i$ are individual specific traits that determine the shape of 
the reaction norm (Fig. \ref{fig:model-struc} A). Variation in $\alpha_i$ 
and $\beta_i$ means individuals can have either uninformative (flat) or 
informative (logistic) norms. Informative reaction norms represent a 
developmental program where the phenotype of the individual is determined by
the environmental conditions under which it grows, thus causing a 
correlation between quality and signal. The size of the signal is constrained 
to take values between zero and one. We assume different values of these traits 
are given by different alleles and are inherited from mother to offspring 
unless, with a small probability ($\mu$), mutation changes the allelic value 
of the offspring by an amount taken from a normal distribution $TN(0,\sigma_\mu)$.

After birth, individuals go through a round of 
viability selection. The survival probability of an individual is 
given by $s_i=1/(1+e^{-k_1-k_2(Q_i-B_i)})$, where $k_1$ and $k_2$ are parameter 
values determining the shape of the survival function. 
Importantly, if $k_2=0$ survival probability is independent of quality, 
while if $k_2>0$ lower quality individuals pay a higher price for similar-sized 
badge, fulfilling the core assumption of the handicap principle 
[@botero_Evolution_2010; @grafen_Biological_1990; @johnstone_Badges_1993].

Individuals who survive engage in a series of pairwise interactions 
where they compete for resources. 
<!-- Individuals in real populations typically  -->
<!-- do not interact at random nor with all individuals in the population  -->
<!-- [@kurvers_Evolutionary_2014]. In order to allow for non-random interactions,  -->
<!-- we use the population vector to define an interaction neighbourhood  -->
<!-- for the focal individual, the size of the neighbourhood ($g$) -->
<!-- is the number of positions around the focal's from which the partner is drawn.  -->
<!-- So, if $g=N$ interactions are global; as $g$ gets smaller  -->
<!-- interactions are more local.  -->
In each interaction individuals must 
decide whether to escalate a fight or not. 
Following the classic *hawk-dove* game [@maynard-smith_Evolution_1982], 
if the focal individual fights and its partner does not, the focal gets 
as pay-off the resource of value $V$ while its partner gets nothing; 
if both individuals restrain  from fighting, then they split up the resource 
in half. If both individuals decide to fight, the winner takes over the resource and the cost of fighting is split among the two players. We further assume that the probability of wining a fight for the focal individual depends on the 
difference in quality between it and its interacting partner; specifically 
it is given by $p_{ij}=1/(1+e^{-k_3(Q_i-Q_j)})$, where $k_3$ is a 
parameter defining how strong the quality difference determines the 
winning probability, and $i$ and $j$ denote the the position 
in the population vector of the focal and its interacting 
partner, respectively.

The decision of whether to escalate a fight against an interacting 
partner can depend on the size of the partner's badge, 
the dependency is determined by the focal's experiences 
through a learning process. We implement learning using 
the actor-critic approach from reinforcement learning (RL) theory [@sutton_Reinforcement_2018;@quinones_Reinforcement_2020; @leimar_Learning_2019;
@leimar_Evolution_2021].
Individuals estimate the reward (pay-off) expected from  interacting 
with partners of different badge sizes (the critic in RL terminology). 
After each interaction they update the estimate of reward proportionally 
to the difference between their current estimate and the observed 
reward (prediction error $\delta$) and to the speed of learning 
($\mathrm{A}$). Furthermore, individuals express different probabilities of 
retreating/attacking depending on the badge size of their opponent 
(the actor in RL). They update the probability of 
retreating/attacking  up or down depending on whether retreating/attacking
leads to an increase in the reward estimation. 
So, if a focal individual decides to escalate a fight against 
an individual with small badge and this leads to an increase in 
the reward estimation, then the focal individual will increase the probability 
of escalating fights with individuals of small badges in the future. 
Given that badge size is a real number between 0 and 1, there are infinitely 
many badge sizes. Thus, the reward estimation, as well as the probability 
of retreating/attacking, must be generalized across different values. 
To implement generalization we use the linear function approximation 
method based on radial basis functions [@sutton_Reinforcement_2018]. 
Specifically, we pick $c$ feature centres, which are evenly spaced values 
along the badge size interval ([0,1]) where the updates are focused. 
For the sake of simplicity we keep the location of these feature centres 
constant and  the same for all individuals, the are stored in the 
vector $\symbf{b}$ . Each of these feature centres 
is associated with a weight for reward estimation and tendency to play 
retreat in a given interaction. The reward estimation and the tendency 
to retreat are calculated (in every interaction) as the sum of the weights 
associated with each feature centre, weighted by the response 
triggered by the feature (Fig. \ref{fig:model-struc} B, dots 
represent the feature weights). The response of each feature centre 
diminishes as a Gaussian function with the distance between the 
feature centre and the badge size of 
the partner (Fig. \ref{fig:model-struc} B, grey line). Formally, the 
reward estimation $\hat{R}$ when the focal individual ($i$) faces 
individual $j$ is given by,
\
\begin{equation}
\hat{R}_{ij} =  \sum_{z=1}^{c} x_z e^{(-\frac{|B_i-b_z|}{2\theta^2})} 
\end{equation}

where $x_z$ is the weight of feature centre $z$ on the reward estimation; 
and $\theta$ is the width of the generalization function. Similarly, 
the tendency to retreat is given by the sum of feature weights associated 
with the actor, and the probability is obtained by applying a logistic 
transformation (Fig. \ref{fig:model-struc} B, black line). 
So formally, the log-odds to retreat when facing individual $j$ is

@@

Individuals interact as the focal individual $n$ times, and the interaction 
partner is chosen at random. Thus, the expected number of interactions for 
each individual is $n+\frac{n(N+1)}{N}$. 
After the interaction round , individuals in the population 
reproduce with a probability proportional to their total pay-off $w_i$, which is a sum of the 
baseline pay-off ($w_0$) and all the pay-offs obtained throughout their 
life. Thus, the combination of natural selection and genetic drift 
changes the distribution of values in $\alpha$ and $\beta$ that segregates 
in the population, effectively changing  the badges expressed and the 
communication system.

```{r model-struc,echo=FALSE,warning=FALSE,fig.width=8,fig.align='center',fig.height=7,fig.show='hold',fig.cap= "Model of communication in the context of aggressive interactions. In A, the reaction norm determines the bib (badge) size, black band on birds chest. Red shows an informative reaction norm, while the blue shows a uninformative reaction norm. B individuals perceiving the signal have a behavioral reaction norm that determines their probability of retreating (black line). The reaction norm arises from generalizing the information from the feature weights (black dots). Feature weaights are values associated with a particular badge size and are updated as individuals interact with each other. The generalization of the information associated with each feature weight is represented by the grey line and its axis, which shows the response triggered by the fourth feature diminishes as the value evaluated is further from the feature center. The learning process moves the feature weights (black dots) up or down depending on whether that leads to an increase in the estimated reward. In C and D, effect of learning on the receiver strategy. Receivers in C face signallers with uninformative reaction norms (blue line in A). The uninformative reaction norm yields badge size of 0.5, thus the behavioural reaction norm in C is only updated around that value.D Receivers face signallers with informative reaction norms (red line in A). Accordingly, receivers develop a threashold-like reaction norm where the decision to retreat increases with the badge size of the interacting partner. Colour scale in C and D indicates the quality of the individual. "}

@@

### Learning under a monomorphic population

>>>>

---
title: |
  The role of learning in status signalling: a modeling approach
author:
- Andrés E. Quiñones:
    email: "andresqp@gmail.com"
    institute: [andes, unine]
    correspondence: true
- Redouan Bshary:
    institute: unine
- C. Daniel Cadena:
    institute: andes
institute:
- andes: Laboratorio de Biología Evolutiva de Vertebrados, Departamento  de Ciencias Biológicas, Universidad de los Andes, Bogotá, Colombia
- unine: Institute of Biology, University of Neuchâtel, Neuchâtel, Switzerland
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
  # word_document: default
  # pdf_document:
    latex_engine: xelatex # default  #
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
fig_caption: yes
fontsize: 12pt
header-includes: |
  ```{=latex}
    \usepackage{float}
    \usepackage{unicode-math}
    \usepackage{setspace}\doublespacing
    \usepackage[backref=page]{hyperref}
    \linespread{2}
    \usepackage{lineno}
    \linenumbers
    \floatplacement{figure}{H}
    \newcommand{\beginsupplement}{ \setcounter{table}{0}         \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
  ```
keep_tex: yes
keywords: My keywords
linespread: 2
bibliography: ../SocialMorphCue.bib
spacing: double
style: C:\Users\andre\Zotero\styles\elife.csl
abstract: |
  Adaptive behavioural responses often depend on qualities of the interacting
  partner of an individual. For example, when competing for resources, an individual
  might be better off escalating fights with individuals of lower quality, while
  restraining  from fighting individuals of higher quality. Communication systems
  involving signals of quality allow individuals to reduce uncertainty regarding the
  fighting ability of their partners and make more adaptive behavioral decisions. 
  However, dishonest individuals can destabilize such communications systems. 
  An open question is whether cognitive mechanisms, such as learning, 
  can maintain the honesty of signals, thus favoring their evolutionary 
  stability. Here we present evolutionary simulations where individuals 
  can produce a signal proportional to their quality and learn along their
  lifetime the best response to the signals emitted by their peers. Our simulations
  replicate previous results where the handicap principle mediates the evolution 
  of signals. As in traditional models without learning, in our simulations 
  learning on the receiver side can mediate the evolution of signals
  of quality on the sender side. When the cost of the signal is proportional to the
  quality of the sender, populations are only composed of honest signalers. In contrast 
  to traditional models which predict the absence of signals when the
  cost is not proportional to the quality of the signaler, our model revealed that
  learning facilitates the evolution of a polymorphism in which populations are
  composed of both honest and dishonest signalers. We argue that learning can 
  be a general cognitive mechanism having a role in a wide range of communication systems 
  and more generally in behavioral responses.
---

@@

The outcome of interactions with conspecific individuals is an important 
determinant of fitness in social animals. Irrespective of whether 
such interactions are cooperative or competitive, the actions of 
interacting individuals partly determine 
their reproductive success. However, the best action for an individual 
engaging in a social interaction might vary depending on its 
own condition, and that of individuals with 
whom it interacts. Thus, acting based on information about interacting 
partners is typically adaptive [@quinones_Negotiation_2016], yet
acquiring such information is far from trivial. 
In some cases, interacting partners (*i.e.* signallers) might be 'willing' to 
provide accurate information, but in others it might be in their 
own interest to conceal information [@johnstone_Recognition_1997] 
or to provide deceptive information [@johnstone_Badges_1993]. 
Typically, in a given context, some individuals benefit 
from broadcasting accurate information, whereas others benefit 
from concealing it. Take, for instance, an interaction between two individuals 
where one can help the other. Given some costs and benefits, 
a potential donor is likely to be interested in helping 
related individuals. Therefore, for a relative of the donor, 
broadcasting kinship would be advantageous, whereas concealing the lack 
of kinship would be better for an unrelated individual. 
Similar scenarios may apply to a variety of aspects of social 
life such as finding mates, feeding offspring, or engaging in
dominance relationships or aggressive interactions 
[@bradbury_Principles_2011; @moller_Female_1988; @tibbetts_Socially_2004].

In dominance and aggressive interactions, a crucial piece of
information to guide individual actions is the fighting ability of interacting 
partners, which is sometimes referred to as Resource Holding Potential 
[RHP, Parker -@parker_Assessment_1974], or simply as quality. 
Responsiveness to the quality of an opponent is central 
to communication systems such as those in which individuals 
exhibit badges of status, where a signal conveys quality 
[@johnstone_Badges_1993; @rohwer_Social_1975a]. 
Such signals must be arbitrary, meaning the signals could potentially be 
produced by high and low quality individuals alike. 
Generally signals can be evolutionarily stable whenever low quality individuals 
pay a higher fitness cost for carrying the signal such that it is not 
in the interest of low quality individuals to fake quality by producing a large badge.
Signal costs inversely proportional to quality may come about as a result of the
production costs of the signal, in which case the signal works like a handicap
[@botero_Evolution_2010; @johnstone_Badges_1993;@grafen_Biological_1990;@zahavi_Mate_1975]. 
Alternatively, the costs can come from 
the aggressive reaction of receivers when the convention of the communication 
is broken [@enquist_Signaling_2010;@tibbetts_Socially_2004]. For example in paper 
wasps, subordinate individuals with experimentally manipulated dominant-like facial
patterns received more aggression from the dominant. In this case, the costs are 
socially mediated rather than developmentally and are triggered by the violation 
of the convention. It is still incluer under this narrative, how the convenvention 
is stablished in the first place.

<!-- should this last part of the paragraph stay ?-->
<!-- An alternative system of communication is one where individuals  -->
<!-- gather and store information about each individual they interact with.  -->
<!-- This information should reliably allow each individual to react adaptively  -->
<!-- to the quality difference with their opponent. This type of  -->
<!-- assessment strategy is often referred to as individual  -->
<!-- recognition [@whitfield_Plumage_1986]. Evidently, individual recognition  -->
<!-- is restricted to  certain interaction structures, because individuals  -->
<!-- are limited by cognitive abilities in how many individuals  -->
<!-- they can recognize and track. Thus, individual recognition and  -->
<!-- badges of status have been assumed to be alternative communication  -->
<!-- systems fulfilling the same function, reducing the costs of  -->
<!-- aggression in contests over resources.  -->

@@

A somewhat ignored component of communication systems in the context 
of aggressive interactions are cognitive aspects of the receiver module. 
Theoretical models often assume that communication systems relying on badges of 
status have reaction norms as their mechanistic underpinning 
[@botero_Evolution_2010], with 
individuals using their own quality and the badges of opponents to determine 
whether to aggressively engage in contests. Reaction norms 
allow individuals to respond to the available information without a large
cognitive burden because the computations involved in the reaction 
norm response do not require much memory nor advanced algorithmic processes. 
This contrasts with other information-processing 
mechanisms like individual recognition. Where individuals associate cues of 
their peers to their quality. Because these 
associations must be learnt throughout the life of individuals, 
they have the usual cognitive requirements of associative learning processes. 
An alternative view of badges of status is that individuals learn 
to react to them based on their experiences [@guilford_Receiver_1991], 
which would imply that receivers must learn to associate signals with 
the fighting ability of bearers just as in systems based on individual 
recognition. In cases where badges of status vary quantitatively 
(*e.g.* in size or intensity), fighting ability may increase monotonically 
with attributes of the badge and would be reinforced by every interaction.
Therefore, in principle, learning in such scenarios would be faster 
than in systems involving individual 
recognition in which the association between signals and their meaning 
varies depending on the interacting partners. In any case, learning may be 
a central cognitive mechanism in both types of communication systems, 
but the role of learning in these contexts has not been thoroughly 
explored in the empirical nor theoretical literature.

@@

Associative learning is a taxonomically widespread cognitive 
mechanism that allows individuals to associate rewards with 
environmental stimuli and thus behave adaptively 
[@heyes_Simple_2012;@macphail_Brain_1982; @staddon_Adaptive_2016; 
@behrens_Associative_2008]. Theory has shown that natural selection 
favours such associations in complex environments where conditions 
are difficult to predict [@dridi_Environmental_2016]. 
Associative learning is a flexible cognitive 
mechanism whose underpinnings show interspecific variation 
[@enquist_Power_2016; @quinones_Reinforcement_2020;@pretot_Factors_2016;
pretot_Comparative_2021]. For example, species can have different learning rates
[@hoedjes_Natural_2010],
some may vary these rates during learning trials which allows them to have more 
flexible learning [@leimar_Flexibility_2023a], and some may include future rewards in their associations [raby_Planning_2007;osvath_Spontaneous_2009].
Associative learning is not a single mechanism, but rather a set of 
cognitive structures and processes that can vary in their scope and complexity. Presumably, these structures and processes have been modified 
by natural selection and could provide new explanations for behavioural 
variation. Nonetheless, associative learning is 
not often included in the narrative of evolutionary explanations of 
behavioural patterns [@fawcett_Exposing_2013; @kamil_Optimal_1983; 
@mcauliffe_Psychology_2015]. Computational models of evolution 
can play an important role to overcome the lack of integration 
between learning and evolution. Reinforcement learning theory 
encompasses a series of computational methods inspired 
on the psychological and neurological mechanisms of associative 
learning [@sutton_Reinforcement_2018]. This set of algorithms allows the 
implementation of biologically realistic problems, 
capturing the essence of learning processes 
[@frankenhuis_Enriching_2018; @quinones_Reinforcement_2020]. 
Furthermore, these algorithms can be embedded in evolutionary simulations to 
generate theoretical predictions of the effect that learning can 
have in behavioural evolution [@leimar_Learning_2019;@leimar_Effects_2022].

Here we present an evolutionary model where individuals use 
associative learning to develop a tendency to behave either aggressively 
or  peacefully in the context of competition over resources, 
depending on a quantitative morphological trait 
(*i.e.* a badge indicating quality) they perceive in their opponents. 
Over evolutionary time, the size of 
the badge evolves as does its dependency on the individual's quality. 
Under this simple set up, individuals can use the badge as a signal of quality. 
We use the model to assess under what conditions 
we expect communication signals to evolve as handicaps or conventions.

@@

We model the evolution of signals indicating individual quality 
in the context of agonistic interactions. 
For simplicity we consider a population of haploid 
individuals with non-overlapping generations. Individuals are born 
every generation with a level of quality ($Q_i$, where $i$ is subscript 
of the population vector of size $N$) given by a number between zero and one, 
which is drawn from a truncated normal distribution 
$N( 0.5,\sigma)$. An individual with quality $0$ has the lowest RHP, 
while an individual with quality $1$ has the highest. As they 
develop, individuals produce a phenotypic signal (badge), 
the size of which depends on their quality according to a reaction norm given by 
$B_i=1/(1+e^{\alpha_i-\beta_i Q_i})$; where $\alpha_i$ and 
$\beta_i$ are individual specific traits that determine the shape of 
the reaction norm (Fig. \ref{fig:model-struc} A). Variation in $\alpha_i$ 
and $\beta_i$ means individuals can have either uninformative (flat) or 
informative (logistic) norms. Informative reaction norms represent a 
developmental program where the phenotype of the individual is determined by
the environmental conditions under which it grows, thus causing a 
correlation between quality and signal. The size of the signal is constrained 
to take values between zero and one. We assume different values of these traits 
are given by different alleles and are inherited from mother to offspring 
unless, with a small probability ($\mu$), mutation changes the allelic value 
of the offspring by an amount drawn from a normal distribution $TN(0,\sigma_\mu)$.

After birth, individuals go through a round of 
viability selection. The survival probability of an individual is 
given by $s_i=1/(1+e^{-k_1-k_2(Q_i-B_i)})$, where $k_1$ and $k_2$ are parameter 
values determining the shape of the survival function. 
Importantly, if $k_2=0$ survival probability is independent of quality, 
while if $k_2>0$ lower-quality individuals pay a higher price for similar-sized 
badges, fulfilling the core assumption of the handicap principle 
[@botero_Evolution_2010; @grafen_Biological_1990; @johnstone_Badges_1993].

Individuals who survive engage in a series of pairwise interactions 
where they compete for resources. 
<!-- Individuals in real populations typically  -->
<!-- do not interact at random nor with all individuals in the population  -->
<!-- [@kurvers_Evolutionary_2014]. In order to allow for non-random interactions,  -->
<!-- we use the population vector to define an interaction neighbourhood  -->
<!-- for the focal individual, the size of the neighbourhood ($g$) -->
<!-- is the number of positions around the focal's from which the partner is drawn.  -->
<!-- So, if $g=N$ interactions are global; as $g$ gets smaller  -->
<!-- interactions are more local.  -->
In each interaction individuals must 
decide whether to escalate a fight or not. 
Following the classic *hawk-dove* game [@maynard-smith_Evolution_1982], 
if the focal individual fights and its partner does not, then the focal gets 
as pay-off the resource of value $V$ while its partner gets nothing; 
if both individuals restrain  from fighting, then they split up the resource 
in half. If both individuals decide to fight, then the winner takes over 
the resource and the cost of fighting is split between the two players. 
We further assume that the probability of wining a fight for the focal 
individual depends on the difference in quality between it and its 
interacting partner; specifically 
it is given by $p_{ij}=1/(1+e^{-k_3(Q_i-Q_j)})$, where $k_3$ is a 
parameter defining how strong the quality difference determines the 
winning probability, and $i$ and $j$ denote the the position 
in the population vector of the focal and its interacting 
partner, respectively.

The decision of whether to escalate a fight against an interacting 
partner can depend on the size of the partner's badge, 
the dependency is determined by the focal's experiences acquired
through a learning process. We implement learning using 
the actor-critic approach from reinforcement learning (RL) theory [@sutton_Reinforcement_2018;@quinones_Reinforcement_2020; @leimar_Learning_2019;
@leimar_Evolution_2021].
Individuals estimate the reward (pay-off) expected from  interacting 
with partners of different badge sizes (the critic in RL terminology). 
After each interaction they update the estimate of reward proportionally 
to the difference between their current estimate and the observed 
reward (prediction error $\delta$) and to the speed of learning 
($\mathrm{A}$). Furthermore, individuals express different probabilities of 
retreating/attacking depending on the badge size of their opponent 
(the actor in RL). They update the probability of 
retreating/attacking depending on whether retreating/attacking
leads to an increase in the reward estimation. 
So, if a focal individual decides to escalate a fight against 
an individual with a small badge and this leads to an increase in 
the reward estimation, then the focal individual will increase the probability 
of escalating fights with individuals of small badges in the future. 
Given that badge size is a real number between 0 and 1, there are infinitely 
many badge sizes. Thus, the reward estimation, as well as the probability 
of retreating/attacking, must be generalized across different values. 
To implement generalization we use the linear function approximation 
method based on radial basis functions [@sutton_Reinforcement_2018]. 
Specifically, we pick $c$ feature centres, which are evenly spaced values 
along the badge size interval ([0,1]) where the updates are focused. 
For the sake of simplicity we keep the location of these feature centres 
constant and  the same for all individuals, the are stored in the 
vector $\symbf{b}$ . Each of these feature centres 
is associated with a weight for reward estimation and tendency to play 
retreat in a given interaction. The reward estimation and the tendency 
to retreat are calculated (in every interaction) as the sum of the weights 
associated with each feature centre, weighted by the response 
triggered by the feature (Fig. \ref{fig:model-struc} B, dots 
represent the feature weights). The response of each feature centre 
diminishes as a Gaussian function with the distance between the 
feature centre and the badge size of 
the partner (Fig. \ref{fig:model-struc} B, grey line). Formally, the 
reward estimation $\hat{R}$ when the focal individual ($i$) faces 
individual $j$ is given by,
\
\begin{equation}
\hat{R}_{ij} =  \sum_{z=1}^{c} x_z e^{(-\frac{|B_i-b_z|}{2\theta^2})} 
\end{equation}

where $x_z$ is the weight of feature centre $z$ on the reward estimation; 
and $\theta$ is the width of the generalization function. Similarly, 
the tendency to retreat is given by the sum of feature weights associated 
with the actor, and the probability is obtained by applying a logistic 
transformation (Fig. \ref{fig:model-struc} B, black line). 
Thus formally, the log-odds to retreat when facing individual $j$ is

@@

Individuals interact as the focal individual $n$ times and the interaction 
partner is chosen at random. Thus, the expected number of interactions for 
each individual is $n+\frac{n(N+1)}{N}$. 
After the interaction round , individuals in the population 
reproduce with a probability proportional to their total pay-off $w_i$, which is a sum of the 
baseline pay-off ($w_0$) and all the pay-offs obtained throughout their 
life. Thus, the combination of natural selection and genetic drift 
changes the distribution of values in $\alpha$ and $\beta$ that segregates 
in the population, effectively changing  the badges expressed and the 
communication system.

```{r model-struc,echo=FALSE,warning=FALSE,fig.width=8,fig.align='center',fig.height=7,fig.show='hold',fig.cap= "Model of communication in the context of aggressive interactions. In A, the reaction norm determines the bib (badge) size, black band on birds chest. Red shows an informative reaction norm, whereas blue shows a uninformative reaction norm. B individuals perceiving the signal have a behavioral reaction norm that determines their probability of retreating (black line). The reaction norm arises from generalizing the information from the feature weights (black dots). Feature weaights are values associated with a particular badge size and are updated as individuals interact with each other. The generalization of the information associated with each feature weight is represented by the grey line and its axis, which shows that the response triggered by the fourth feature diminishes as the value evaluated is further from the feature center. The learning process moves the feature weights (black dots) up or down depending on whether this leads to an increase in the estimated reward. In C and D, effect of learning on the receiver strategy. Receivers in C face signallers with uninformative reaction norms (blue line in A). The uninformative reaction norm yields a badge size of 0.5, thus the behavioural reaction norm in C is only updated around that value. D Receivers face signallers with informative reaction norms (red line in A). Accordingly, receivers develop a threashold-like reaction norm where the decision to retreat increases with the badge size of the interacting partner. Colour scale in C and D indicates the quality of the individual. "}

@@

### Learning in a monomorphic population
